Utility of Diffusion Models for Predicting and Diagnosing the Spread of Conservation Behaviors

### 2021-11-20 DOI_Review/Manuscript.rmd
Text edits through quantitative applications

date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Community-based conservation, diffusion of innovations, cultural evolution, complex systems
bibliography: [./DOI_Conservation_Review.bib]
abstract: A common goal of conservation practitioners is to design interventions that benefit both people and the environment, and as a result are adopted robustly by local communities. Diffusion theory provides generalized heuristics and models for how learned behaviors spread within a population and has been increasingly identified by researchers as useful for understanding conservation adoption. Current prescriptions have nevertheless been largely ambiguous as to the practical applications of diffusion theory to conservation work, and in particular, in coupled human-natural systems. Here, we demonstrate the capacity for a process-based diffusion model to reasonably predict the long-term rates of conservation adoption under tightly coupled behavior-environment feedbacks. Data on conservation adoption are generated through an agent-based simulation which highlights our dynamics of interest. We caution however, that while diffusion models may predict the pattern with which a conservation intervention will spread through a population, they give little diagnostic insight as to why. We show that the social learning mechanism employed by a population can alter the final proportion of individuals that adopt an intervention, and argue that conservation researchers and practitioners can leverage theory from social learning and cultural evolution to better design and diagnose projects. Lastly, we emphasize that developing and fitting process models in statistical frameworks traditionally reserved for linear regression alone, may pose a productive new avenue for conservation science, particularly in concert with simulation techniques and insights from complex systems.

@@

# Diffusion Theory
Promoting the widespread adoption of some novel (to the individual) behavior is not an issue unique to community-based conservation; producers of cigarettes and snack foods, as well as public health professionals, have been grappling with these questions for quite some time. Diffusion theory recognizes that this process is primarily governed by the level of uncertainty an individual has about a novel behavior and their personal threshold for the amount of certainty required to make a decision [@rogers_diffusion_2010]. When an individual is introduced to a new behavior, they often have a high degree of uncertainty as to the risks and benefits of adopting and will wait to adopt until that uncertainty is reduced to a tolerable level. The primary way individuals reduce uncertainty about the outcome of adopting a particular behavior is through second-hand observation of early adopters. Therefore, the core of diffusion theory is the notion that individuals imitate those around them when making decisions about adopting a novel behavior or not, rather than weighing the pros and cons independently.

This assumption about human decision-making has been validated repeatedly in cultures around the world; individuals do not act as cost-benefit analysts, in which they adapt their behaviors based on the environment alone [@henrich_search_2001]. Our dependence on social learning to make decisions has profound and specific implications for patterns of behavioral adoption at the population level [@smaldino_adoption_2017]. We observe these patterns through diffusion curves which show the proportion of a population that has adopted a behavior as a function of time [@hoppitt_effect_2010; @rogers_diffusion_2010; @ryan_acceptance_1950].

Behavioral spread through imitation produces a sigmoidal “S” shaped diffusion curve, which is observed in the vast majority of empirical studies on behavioral adoption [@henrich_cultural_2001]. This “S” shaped curve is produced regardless of the social learning mechanism employed by a population and can even be observed when a portion of the population relies on individual learning [@henrich_cultural_2001]. Behavioral adoption thus exhibits equifinality, where many different underlying mechanisms may produce similar or even identical patterns over time [@barrett_equifinality_2019]. Critically, this makes it virtually impossible to infer the learning mechanism(s) used by a population from the pattern of adoption over time alone.

# Diffusion Theory in Conservation

Individuals tend to place more weight on imitation when making decisions in which the outcomes can not be easily predicted independently [@perreault_bayesian_2012]. This has been explicitly demonstrated under variable, rapidly changing environments [@boyd_culture_1988; @richerson_not_2008]. It follows then, that the adoption of novel conservation behaviors by a community is likely to be achieved through social imitation (i.e. diffusion), especially given worsening climatic instability and regime changes resulting from widespread biodiversity loss.

Indeed, the sigmoidal pattern of behavioral adoption characteristic to diffusion has been observed in over 80% of case studies on community-based conservation interventions worldwide [@mascia_when_2018; @mills_how_2019]. Conversely, in contexts where communication and thus imitation are limited, such as across island chains, the adoption of community-based conservation interventions is considerably attenuated [@pietri_information_2009-1]. With the recognition that the patterns of uptake for community-based conservation practices are directly linked to behavioral imitation, researchers have begun suggesting diffusion theory as a lens for understanding and implementing individual interventions [@abernethy_two_2014; @mbaru_key_2017; @niemiec_efficacy-based_2021].

Meta-level studies have recently come to the same conclusion. For example, in a review of 22 conservation interventions referenced above, the majority of initiatives were adopted by less than 30% of potential adopters, leading the authors to point to diffusion theory as an avenue for increasing the proportion of adopters [@mills_how_2019]. And a recent framework for understanding the establishment, persistence, and spread of community‐based conservation practices further identified diffusion theory as useful for guiding the development of conservation interventions [@mahajan_theorybased_2020].

These studies are compelling in their conclusion that diffusion theory can be productively applied to the spread of community-based conservation interventions, but ambiguous in exactly how someone might practically apply this insight. We envision a practitioner who is monitoring the weekly, monthly, or even yearly uptake rates of an intervention and is hoping to predict if and when it will “take off” in the population, what the final proportion of adopters will be, and/or what they can do to promote its spread. The aim of this review is to explicitly demonstrate how diffusion theory can be used to generate predictions about the patterns of uptake of conservation behaviors given realistic data, and the limitations of this theory as a diagnostic tool and for application to complex, coupled human-natural systems.

>>>>

date: "`r format(Sys.time(), '%B %d, %Y')`"
keywords: Community-based conservation, diffusion of innovations, cultural evolution, complex systems
bibliography: [./DOI_Conservation_Review.bib]
abstract: A common goal of conservation practitioners is to design interventions that benefit both people and the environment, and as a result are adopted robustly by local communities. Diffusion theory provides generalized heuristics and models for how learned behaviors spread within a population and has been increasingly identified by researchers as useful for understanding conservation adoption. Here, we elaborate on the need for better understanding the spread of conservation behaviors, and review diffusion theory and its past application to these problems. We follow this with a practical demonstration of how the mathematical models which fit the diffusion process can be used to diagnose and predict the uptake of conservation behaviors even under tightly coupled behavior-environment feedbacks. Data on conservation adoption are generated through an agent-based simulation which highlights our dynamics of interest. Throughout this review, situate diffusion theory in the broader literature on cultural evolution and use these insights to refine our claims about how diffusion can be productively applied to conservation. Specifically, we show that the social learning mechanism individuals employ can alter the population pattern of adoption and the final proportion of individuals that adopt an intervention. Overally, we provide a quantitative approach for modeling the spread of conservation behaviors and argue for a more complete incorporation of cultural evolutionary theory when considering diffusion in these systems.

@@

# Diffusion Theory
Promoting the widespread adoption of some novel (to the individual) behavior is not an issue unique to community-based conservation; producers of cigarettes and snack foods, as well as public health professionals, have been grappling with these questions for quite some time. Diffusion theory recognizes that this process is primarily governed by the level of uncertainty an individual has about a novel behavior and their personal threshold for the amount of certainty required to make a decision [@rogers_diffusion_2010]. When an individual is introduced to a new behavior, they often have a high degree of uncertainty as to the risks and benefits of adopting and will wait to adopt until that uncertainty is reduced to a tolerable level. The primary way individuals reduce uncertainty about the outcome of adopting a particular behavior is through second-hand observation of early adopters. Therefore, at the core of diffusion theory is the observation that individuals imitate those around them when making decisions about adopting a novel behavior or not, rather than weighing the pros and cons independently [@henrich_search_2001].

When modifying their beliefs or behaviors, people do not imitate any other individual at random [@bandura_social_1977]. Instead, they preferentially adopt traits by employing a variety of learning biases such as success-biased learning where individuals who are perceived as more successful are more likely to be copied than unsuccessful individuals [@boyd_culture_1988; @laland_social_2004]. As a result, the cumulative culture of a population evolves over time as various beliefs and behaviors are unequally copied and their frequency within a population changes. While a full review of cultural evolution is beyond the scope of this paper, the general intuition is that populations carry a variable pool of information which is shaped by events in the lives of individuals such as social observation or natural selection [@mesoudi_cultural_2016]. This information affects individuals’ behaviors, which in-turn may affect their environments, and further alter which traits are adopted in the future [@durham_coevolution_1991]. Thus, cultural evolution provides an intuitive way to scale from individual learning and decision making to population level patterns and even population level behavior-environment feedbacks.

Diffusion of novel beliefs and behaviors is a specific instance of cultural evolution, where the trait being transmitted is new to the population and outcompetes some previous cultural variant, such as in the classic example of hybrid corn seed adoption in Iowa farmers [@mesoudi_cultural_2016; @ryan_acceptance_1950]. We observe these patterns through diffusion curves which show the proportion of a population that has adopted a belief or behavior on the Y-axis, as a function of time on the X-axis [@hoppitt_effect_2010; @rogers_diffusion_2010; @ryan_acceptance_1950]. The exact shape of these diffusion curves is highly dependent on the learning mechanism employed by individuals in the population, as well as contextual factors like the population density or ease of trialing the cultural variant [@smaldino_adoption_2017; @henrich_cultural_2001; @rogers_diffusion_2010 ].

In the vast majority of empirical studies, the diffusion of a novel trait produces a sigmoidal “S” shape, where it initially spreads slowly, followed by a period of rapid adoption, then again spreads slowly until it reaches fixation [@rogers_diffusion_2010 ]. This “S” shaped curve is indicative of biased social learning, in contrast to the relatively rare “R” shaped curves observed when individuals learn through independent trial and error (fig 1) [@henrich_cultural_2001]. These two curve shapes however are not completely discrete, as the combination of learning strategies employed by individuals in a population shifts more toward individual or social learning, we expect to see a shift toward an “R” shape or “S” shape respectively [@henrich_cultural_2001].

![Canonical "S" and "R" diffusion curves. "R" curves are generally indicative of a higher proportion of individuals employing independent learning. "S" curves are generally indicative of a higher proportion of individuals employing social learning. ](./SvsR2.png)

# Diffusion Theory in Community-Based Conservation

Just as the sigmoidal pattern indicative of social learning is seen in the majority of cases on behavioral adoption broadly, scholars have also noticed this pattern in over 80% of case studies on the uptake of community-based conservation interventions worldwide [@mascia_when_2018; @mills_how_2019]. High-level studies have begun identifying diffusion theory as potentially useful for diagnosing why some community-based conservation interventions spread and persist, while others fizzle out. For example, in a review of 22 conservation interventions referenced above, the majority of initiatives were adopted by less than 30% of potential adopters. The authors point to a sigmoidal diffusion pattern to explain the slow start of many initiatives and suggest that practitioners may actively spread information on interventions (as in Abernethy et al. [-@abernethy_two_2014]), or facilitate communication to increase adoption rates [@mills_how_2019; @greenhalgh_diffusion_2004]. Further, a recent framework for understanding the establishment, persistence, and spread of community‐based conservation practices identifies qualitative insights, such as intervention observability, from diffusion theory as useful in guiding the development of conservation interventions [@mahajan_theorybased_2020].

As these high-level studies suggest, qualitative insights from diffusion theory have been productively applied to understand the rate and success of spread of conservation behaviors in single case studies around the world. In Kenya for example, Mbaru and Barnes [-@mbaru_key_2017] show that identifying socially well-connected individuals can facilitate the rapid diffusion of information on conservation objectives. Similarly, in Fort Collins, CO, researchers demonstrate that both the speed and extent of diffusion of information on native plant gardening likely increase with after motivating communication in the population [@niemiec_efficacy-based_2021]. Romero-de-Diego et al. [-@romero-de-diego_drivers_2021] use expert interviews to identify that observability of benefits was a key driver of the rapid scaling-up of wildlife management areas in Mexico. Conversely, in contexts where communication and thus imitation are limited, such as across island chains, the rate and degree of spread of community-based conservation interventions is considerably attenuated [@pietri_information_2009-1]. This finding mirrors those of studies which look at more general cultural adoption across island chains, further suggesting that conservation behaviors may spread similarly to technological or ideological innovations [@kline_population_2010; @beheim_inheritance_2011].

These studies are compelling in their conclusion that diffusion theory can be productively applied to the spread of community-based conservation, but have been critiqued as qualitative suggestions may miss important contextual factors unique to each intervention [@steenbergen_theory_2021-1; @lam_scaling_2020]. Below, we make a quantitative argument for using diffusion theory to understand the spread of community-based conservation interventions. We envision a practitioner who is monitoring the weekly, monthly, or even yearly uptake rates of an intervention and is hoping to predict if and when it will “take off” in the population, what the final proportion of adopters will be, and/or what they can do to promote its spread. The aim of this review is to explicitly demonstrate how diffusion theory can be used to generate predictions about the patterns of uptake of conservation behaviors given realistic data, and diagnose when in the lifetime of an intervention individuals may be dropping out.
